Trends and Challenges in Toronto’s Homeless Population (2018-2024)

### 2024-09-25 paper/paper.qmd
cleaning data（remove some information)
add sketcht

---
title: "Trends and Challenges in Toronto’s Homeless Population (2018-2024)"
subtitle: "A Focus on Demographics and Chronic Homelessness"
author: 
  - Jianing Li
thanks: "Code and data are available at: https://github.com/JianingLi1225/Toronto_Homeless"
date: today
date-format: long
abstract: "This paper explores Toronto’s homeless population from 2018 to 2024, focusing on age, gender, and chronic homelessness trends. The data reveal a decline during the pandemic, followed by a rebound, with males and individuals aged 35-44 making up the majority. The number of chronically homeless also increased significantly. These findings highlight the need for targeted interventions to address chronic homelessness and improve housing policies."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In this essay, the (@sec-data-overview) section explains the background and necessity of using this dataset, followed by an explanation of the variables mentioned and the methods used. The (@sec-data-results) section shows the changes in the number of homeless population and the proportion of each age group, gender and chronic homelessness. Graphs and related explanations are provided. The (@sec-discussion) section further analyses the information presented in the graphs.

@@

# Loading the Cleaned Dataset #
cleaned_shelter_data <- read_csv(here::here("data", "analysis_data", "shelter_analysis_data.csv"))

@@

# This code is based on knowledge from Alexander (2023).
# Load ggplot2 for plotting
ggplot(total_population_data, aes(x = Date, y = Homeless_Count)) +
  geom_line(color = "blue", linewidth = 1.5) +  
  labs(x = "Date",
       y = "Homeless Population Count") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Create a pie chart with percentage labels
# This code is based on knowledge from Alexander (2023) and further refined using the RStudio Help documentation.
ggplot(total_data_long, aes(x = "", y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4) +
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-gender-distribution
#| fig-cap: "Homeless Population by Gender in Toronto (2018-2024)"
#| 
# Summarize data by year for different genders
# This logic is similar to the previous graph of age group 
gender_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(Male = sum(Male_Count, na.rm = TRUE),
            Female = sum(Female_Count, na.rm = TRUE),
            Other = sum(Transgender_Non_Binary_Count, na.rm = TRUE))

# Calculate total population for each gender over the entire period
total_population <- gender_data %>%
  summarise(Male = sum(Male), Female = sum(Female), Other = sum(Other))

# Reshape data for plotting
total_population_long <- pivot_longer(total_population, 
                                      cols = c("Other", "Female", "Male"), 
                                      names_to = "Gender", 
                                      values_to = "Total_Count")

@@

# Create a pie chart similar to the one used for age groups
# This code is based on knowledge from Alexander (2023) and further refined using the RStudio Help documentation.
ggplot(total_population_long, aes(x = "", y = Total_Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4, fontface = "bold", color = "gray30") +  
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_manual(values = c("Male" = "#7FB3D5", "Female" = "#F5B7B1", "Other" = "#D2B4DE"))

@@

## Overview of Homeless Trends (2018-2024)
In @sec-data-results，trends in Toronto's homeless population from 2018 to 2024 are presented, focusing on key factors such as age, gender, and chronic versus non-chronic conditions. The homeless population fluctuates significantly, being significantly impacted in specific years such as 2020 and 2021[@fig-total-population], with males and the 35-44 age group making up the majority of the homeless population[@fig-age-distribution,@fig-gender-distribution], and chronic homelessness showing an increasing trend[@fig-chronic-population].
## Impact of COVID-19 on Homelessness
The homeless population declined significantly in 2020 and 2021, likely due to COVID-19. To meet social distancing requirements, the Toronto government reduced shelter capacity and opened temporary shelters, including additional spaces in hotels and motels[@may2023scan]. These actions may explain the drop during these years. However, the rebound in homelessness afterward suggests that issues like housing affordability and lack of social support remain unresolved, highlighting the need for sustainable long-term solutions.
## Demographic and Chronic Homelessness Insights
An analysis of age, gender, and chronic homelessness provides key insights into Toronto's homeless population. Homelessness is concentrated among working-age adults, especially those aged 35-44 and 25-34, likely due to challenges with employment or housing stability. In terms of gender, males consistently make up the majority, while non-binary individuals account for only 1.4%. This suggests shelters should consider different designs for men and women, while also addressing the needs of non-binary individuals, despite their smaller numbers. The rise in chronic homelessness suggests that existing measures are not enough to help people exit homelessness permanently.

>>>>

---
title: "Trends and Challenges in Toronto’s Homeless Population (2018-2024)"
subtitle: "A Focus on Demographics and Chronic Homelessness"
author: 
  - Jianing Li
thanks: "Code and data are available at: https://github.com/JianingLi1225/Toronto_Homeless"
date: today
date-format: long
abstract: "This paper explores Toronto’s homeless population from 2018 to 2024, focusing on age, gender, and chronic homelessness trends. The data reveal a decline during the pandemic, followed by a rebound, with males and individuals aged 35-44 making up the majority. The number of chronically homeless also increased significantly. These findings highlight the need for targeted interventions to address chronic homelessness and improve housing policies."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

In this essay, @sec-data-overview explains the background and necessity of using this dataset, followed by an explanation of the variables mentioned and the methods used. @sec-data-results shows the changes in the number of homeless population and the proportion of each age group, gender and chronic homelessness. Graphs and related explanations are provided. @sec-discussion further analyses the information presented in the graphs.

@@

# Loading the Cleaned Dataset #
cleaned_shelter_data <- read_csv(here::here("data", "analysis_data", "analysis_data.csv"))

@@

# This code draws on concepts from Alexander (2023)
# Load ggplot2 for plotting
ggplot(total_population_data, aes(x = Date, y = Homeless_Count)) +
  geom_line(color = "blue", linewidth = 1.5) +  
  labs(x = "Date",
       y = "Homeless Population Count") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

# Create a pie chart with percentage labels
# This code draws on concepts from Alexander (2023)
ggplot(total_data_long, aes(x = "", y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4) +
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-gender-distribution
#| fig-cap: "Homeless Population by Gender in Toronto (2018-2024)"
#| 
# Summarize data by year for different genders
# This logic is similar to the previous graph of age group 
gender_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(Male = sum(Male_Count, na.rm = TRUE),
            Female = sum(Female_Count, na.rm = TRUE),
            Transgender_Non_Binary = sum(Transgender_Non_Binary_Count, na.rm = TRUE))  # Change 'Other' to 'Transgender/Non-binary'

# Calculate total population for each gender over the entire period
total_population <- gender_data %>%
  summarise(Male = sum(Male), Female = sum(Female), Transgender_Non_Binary = sum(Transgender_Non_Binary))  # Adjust for renamed column

# Reshape data for plotting and update Transgender_Non_Binary to Transgender/Non-binary
total_population_long <- pivot_longer(total_population, 
                                      cols = c("Transgender_Non_Binary", "Female", "Male"), 
                                      names_to = "Gender", 
                                      values_to = "Total_Count") %>%
  mutate(Gender = recode(Gender, "Transgender_Non_Binary" = "Transgender/Non-binary"))  # Replace underscore with /

@@

# Create a pie chart similar to the one used for age groups
# This code is based on knowledge from Alexander (2023) and further refined using the RStudio Help documentation.
ggplot(total_population_long, aes(x = "", y = Total_Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4, fontface = "bold", color = "gray30") +  
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_manual(values = c("Male" = "#7FB3D5", "Female" = "#F5B7B1", "Transgender/Non-binary" = "#D2B4DE"))  
```

@@

## Overview of Homeless Trends (2018-2024)
In @sec-data-results，trends in Toronto's homeless population from 2018 to 2024 are presented, focusing on key factors such as age, gender, and chronic versus non-chronic conditions. The homeless population fluctuates significantly, being significantly impacted in specific years such as 2020 and 2021[@fig-total-population], with males and the 35-44 age group making up the majority of the homeless population[@fig-age-distribution,@fig-gender-distribution], and chronic homelessness showing an increasing trend[@fig-chronic-population].

## Impact of COVID-19 on Homelessness
The homeless population declined significantly in 2020 and 2021, likely due to COVID-19. To meet social distancing requirements, the Toronto government reduced shelter capacity and opened temporary shelters, including additional spaces in hotels and motels[@may2023scan]. These actions may explain the drop during these years. However, the rebound in homelessness afterward suggests that issues like housing affordability and lack of social support remain unresolved, highlighting the need for sustainable long-term solutions.

## Demographic and Chronic Homelessness Insights
An analysis of age, gender, and chronic homelessness provides key insights into Toronto's homeless population. Homelessness is concentrated among working-age adults, especially those aged 35-44 and 25-34, likely due to challenges with employment or housing stability. In terms of gender, males consistently make up the majority, while non-binary individuals account for only 1.4%. This suggests shelters should consider different designs for men and women, while also addressing the needs of non-binary individuals, despite their smaller numbers. The rise in chronic homelessness suggests that existing measures are not enough to help people exit homelessness permanently.
